---
title: "Analysis of gene expression data from qPCR in R"
author: ""
date: "`r format(Sys.time(), '%B, %Y')`"
output:
  bookdown::html_document2:
    fig_caption: TRUE # render html with figure captions
    number_section: TRUE # if you want number plots
    toc: TRUE # table of content true
    toc_depth: 3 # upto three depths of headings (specified by #, ## and ###)
    toc_float: 
      colapsed: FALSE
      smooth_scroll: FALSE
   # highlight: tango # specifies the syntax highlighting style
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

## What is this script for?  

This is a simple R Markdown script for researchers interested in transitioning their qPCR analysis from GUI based platforms such as Prism or Excel to R. All files required to run the analysis with an example data set can be found [here](https://github.com/fabianstamp/qPCR-data-analysis).

## How to start?    

You can run the analysis as is with the provided files. They include this script (**qPCR-data-analysis-script.Rmd**) and an example data set (**qPCR_data.xlsx**).  If you want to run the analysis with your own experimental data, you will required the following information:  

 * A list matching each tested sample with a condition (**condition**)
 * Positional information of replicates on the plate (**replicates**)
 * Positional information of samples on the plate (**samples**)
 * Positional information of genes tested on the plate (**genes**)  
 * Raw data from the light cycler containing the Ct values per well (**data**)
 
The example data was acquired using QuantStudio 7 (384 well plate). When adapting this script, make sure to work with the same format (384 well). In case of using a different format (e.g. 96 well), you will have to change the import options accordingly.

## Analysis details 

The workflow consists of 3 steps:  

1. **Data import and formatting**: for easy implementation, the experimental data is imported directly from an Excel file containing the 5 sheets: *condition*, *replicates*, *samples*, *genes* and *data*.  
  
2. **Data analysis**: this includes (**A**) calculation of normalized gene expression, (**B**) calculation of ∆∆Ct and Log2FoldChange per sample and (**C**) selection of adequate statistics using a decision tree.  

3. **Plotting**: visualization of data using ggplot2 with the results from the statistical tests indicating group differences.


# Setup  

## Load libraries
```{r libraries, message=FALSE}
library(readxl) # readxl is used to import data from excel (.xlsx)
library(tidyverse) # data wrangling
library(ggpubr) # required for stat_compare_means() and stat_pvalue_manual()
library(rstatix)  # https://github.com/kassambara/rstatix
library(stats) # required for padjust()
library(car) # required for leveneTest()
library(kableExtra) # scrollable tables in rendered html report
```

## Set global variables  

**Set ID, subsets and reference genes**

Manually set the experimental "ID" and "Details". These will be used for labeling results and plots.

```{r set IDs}
ID = "Experiment ID"
Details = "Brain WT vs KO"
```
**Define levels of the experimental condition**  
Set the order of levels of your experimental conditions. This will determine in which order the results are plotted.
```{r set condition levels}
condition_levels <- c("WT", "KO") #set order of levels
```

**Define reference genes**  
Define which genes will be used as a reference for normalization of gene expression.
```{r set HKs}
HK1 = "reference_gene_1" #reference gene 1
HK2 = "reference_gene_2" #reference gene 2
HK <- c("reference_gene_1","reference_gene_2")  # vector of both reference genes
```

#  Load data

**Set path to files**  
```{r set path to files}
# Path to the qPCR_data.xlsx file
path <- file.path("..", "data", "qPCR_data.xlsx")
```


## Load Ct values from light cycler

**Note:** In case your light cycler results table contains a large header, you can skip these rows by using the skip argument.
```{r load cycler data, message=FALSE, warning=FALSE}
data <-
  read_excel(path, sheet = "data", skip = 0) %>%  # import sheet "data"
  dplyr::select("Well Position", "CT") %>%  # only select columns required for the analysis
  mutate(CT = as.numeric(CT)) %>%  # set Ct as numeric
  mutate(CT = replace(CT, is.na(CT), "")) # replace NA values 
```


## Load gene list

Import the positional information of which gene was tested in each well. 
```{r load genes}
genes <-
  read_excel(path, sheet = "genes", col_names = paste0("V", 1:24)) %>% # import sheet "genes" and add provisional column headers
  mutate(row = LETTERS[1:16]) %>% # 16 rows: add the Letters to create positional information
  pivot_longer(cols = c(1:24)) %>% # 24 columns
  mutate(name = substring(name, 2)) %>% # This will create a positional coding, replacing the name "V" with a number
  mutate("Well Position" = paste(row, name, sep = ""), # Create a new ""Well Position"" column by pasting the values of the row and name columns together
         Gene = value) %>% # Create a "Gene" column, that contains values
  select("Well Position", Gene) %>%  # Select columns
  replace(is.na(.), "") %>%  # replace NA values
  dplyr::mutate(Gene = factor(Gene))  # convert Gene to factor
#head(genes)
```

## Load condition information  
Information matching sample numbers with the corresponding condition (e.g. Sample 1 = WT, Sample 2 = KO).
```{r load condition}
condition  <-
  read_excel(path,
             sheet = "condition",
             col_types = "text",
             col_names = TRUE) %>%   # import sheet "condition"
  mutate(Condition = factor(Condition, levels = condition_levels)) # define "Condition" as factor and set levels
#head(condition)
```

## Load sample information

Load sheet containing positional information of samples, i.e. which sample was measured in which well.
```{r load samples}
samples <-
  read_excel(
    path,
    sheet = "samples",
    col_types = "text",
    col_names = paste0("V", 1:24)
  ) %>% # import sheet "samples" and add provisional column headers
  mutate(row = LETTERS[1:16]) %>% # 16 rows: add the Letters to create positional information
  pivot_longer(cols = c(1:24)) %>% # 24 columns
  mutate(name = substring(name, 2)) %>% # This will create a positional coding, replacing the name "V" with a number
  mutate("Well Position" = paste(row, name, sep = ""),
         # Create a new ""Well Position"" column by pasting the values of the row and name columns together
         Sample = value) %>%  # Create a "Sample" column, that contains values
  select("Well Position", Sample) %>%   # Select columns
  replace(is.na(.), "") %>%  # replace NA values
  dplyr::mutate(Sample = factor(Sample))  # convert "Sample" to factor
#head(samples)
```



## Load replicates information  

Load sheet containing positional information of replicates, i.e. which replicate was measured in which well. Usually you will have 2-3 replicates per sample and gene.

```{r load reolicates}
replicates <-
  read_excel(
    path,
    sheet = "replicates",
    col_types = "text",
    col_names = paste0("V", 1:24)
  ) %>% # import sheet "replicates" and add provisional column headers
  mutate(row = LETTERS[1:16]) %>% # 16 rows: add the Letters to create positional information
  pivot_longer(cols = c(1:24)) %>% # 24 columns.
  mutate(name = substring(name, 2)) %>%  # This will create a positional coding, replacing the name "V" with a number
  mutate("Well Position" = paste(row, name, sep = ""),
         # Create a new ""Well Position"" column by pasting the values of the row and name columns together
         Replicates = value) %>% # Create a "Replicate" column, that contains the  "value" values
  select("Well Position", Replicates) %>%
  replace(is.na(.), "") %>%  # replace NA values
  dplyr::mutate(Replicates = factor(Replicates))  # convert "Replicates" to factor
#head(replicates, n = 3)
```


## Merge data frames  

Merge the **condition**, **replicates**, **samples**, **genes** and **data** tibbles into one data frame (df). "NTP" is used here to refer to "no template" containing control wells that are not used for the expression analysis.
  
```{r merge data frames}
df <- merge(data, samples, by="Well Position") %>% 
  merge(genes, by="Well Position") %>% 
  merge(replicates, by = "Well Position") %>% 
  merge(condition, by = "Sample") %>% 
  filter(!CT == "Undetermined") %>%  # remove all undetermined wells
  mutate(CT = as.numeric(CT)) %>% # define CT as numeric
  filter(Sample != "NTP")  # drop NTP wells
#head(df, n = 3)
```



# Analysis  

## Plot raw Ct values

Plot raw Ct values to identify outliers that result from technical issues (e.g. pippeting errors).  

```{r plot raw Ct,fig.height = 10, fig.width = 15, fig.align = "center"}
Ct_plot <- df %>%
  ggplot(aes(Sample, CT, fill = Gene)) +
  geom_boxplot(alpha = 0.5) +
  geom_point(aes(fill = Gene), shape = 21,
             size = 3) +
  facet_wrap(Gene ~ Condition,
             scales = "free") + # split plot
  labs(x = "Sample", y = "CT") + # axis labels
  ggtitle("CT values") + # title
  theme(plot.title = element_text(hjust = 0.5)) + # center title
  theme(legend.position = "right") + # position of legend
  theme_classic() + # add classic theme
  theme(plot.title = element_text(size = 20, face = "bold")) # increase title size

# save plot
ggsave(
  "Ct_plot.pdf",
  plot = Ct_plot,
  device = "pdf",
  path = "../output/plots",
  scale = 1,
  height = 20,
  width = 30,
  units = "cm",
  dpi = 300,
  limitsize = TRUE,
)
# print to screen
print(Ct_plot)
```


## Calculation of normalized gene expression and Log2FoldChange  

**Steps:**  

**1.** Calculate mean CT values across technical replicates per sample and gene.  

**2.** Calculate the geometric mean for the reference genes (HKs) you want to normalize the expression to.  

**3.** Calculate DCT per sample: ∆Ct = Ct (gene of interest) – geometric mean of Ct (reference genes).  

**4.** Calculate DDCT per sample: ∆∆Ct = ∆Ct (KO sample) – ∆Ct (WT samples average).  

**5.** Calculate the relative fold gene expression (FoldChange) per sample: Fold gene expression = 2^-(∆∆Ct).  

**6.** Transform LogFoldChange to log2 scale(Log2FoldChange) for statistical testing.  





### (1) Calculate mean CT values acrtoss techical replicates per sample and gene  

```{r calculate mean Ct, message=FALSE}
df1 <- df %>%
  select(Sample, Gene, CT, Condition) %>% # select columns of interest
  group_by(Gene, Sample, Condition) %>% # group values by Gene, Sample and Condition
  summarize(CT = mean(CT, na.rm = T)) %>% # use the summarize function to calculate the mean CT - overwriting the CT column. Remove NA values, since it stops the function from working
  pivot_wider(
    id_cols = c(Sample, Condition),
    names_from = Gene,
    values_from = CT
  ) # pivot to wide format
```
### (2) Calculate the geometric mean expression of reference genes

```{r calculate geometric mean of HKs, message=FALSE}
df2 <- df1 %>%
  gather(Gene, CT_mean,-c(Sample, Condition, all_of(HK)))  # convert to long format

# Calculate the geometric mean for reference genes
geo_mean <- df2 %>%  select(all_of(HK)) %>%
  gather(key = Ref, value = value,-Sample) %>%
  group_by(Sample) %>%
  summarise(Geo_HK = exp(mean(log(value))))

# join df2 with geo_mean
df3 <- df2 %>% 
inner_join(geo_mean, by = "Sample")
```

### (3-6) Calculate ∆Ct, ∆∆Ct and Log2FoldChange  

```{r Log2FoldChange}
df3 <- df3 %>% 
mutate(DCT = CT_mean - Geo_HK) %>%  # calculate delta CT by subtracting the geometric mean of the house keeper genes from the individual sample CT means
  ungroup() %>% 
  group_by(Gene, Condition) %>% # group by Gene and Condition 
  mutate(DCT_mean_WT = mean(DCT)) %>%  # calculate DCT for each Condition. There might be a more elegant version to do this only for the WT animals (?)
  ungroup() %>% 
  group_by(Gene) %>% 
  mutate(DDCT = DCT - DCT_mean_WT[Condition=="WT"]) %>% # subtract the DCT of the WT animals from the DCT of the individual samples (treated sample - untreated sample)
  mutate(FoldChange = 2^-(DDCT)) %>% # calculate FoldChange 
  mutate(Log2FoldChange = log2(FoldChange)) # Transform onto log2 scale
```


### Optional: Calculate N number for labeling plots  

Determine the N per condition and create a label to add to plots
```{r N for plotting}
df3 <-df3 %>% 
  group_by(Condition, Gene) %>% 
   mutate( N = n()) %>% # add N number
  mutate(Condition_label = paste0(Condition,"\n","(N=",N,")")) %>% # add annotation for plots
  mutate(Condition_label = as.factor(Condition_label)) %>%  # define annotation as factor
  mutate(Experiment = as.factor(ID)) %>% 
  ungroup()
#head(df3)
```


## Statistical analysis  

Gene expression values are often non normally distributed and heavily skewed. Specifically, down-regulated genes have a value between 0 and 1 (1 = no expression change), while up-regulated genes can have a value between 1 and positive infinity. By log transforming the gene expression ratios, they become symmetrical distributed around 0. Therefore, for the statistical analysis you should log-transform the expression data with the assumption that gene regulation and protein expression follow a log-normal distrubution.

### Calculate summary statistics per condition  

```{r summary statistics, message=FALSE}
#subset data frame
df_stats <- df3 %>%
  select("Condition", "Log2FoldChange", "Gene")

df_sum <- df_stats %>%
  group_by(Condition, Gene) %>%   # group
  dplyr::summarise(
    n = n(),
    # how many observations
    mean = round(mean(Log2FoldChange), digits = 2),
    # calculate mean
    sd = round(sd(Log2FoldChange), digits = 2),
    # calculate sd
    p_shapiro = round(shapiro.test(Log2FoldChange)$p.value, digits =
                        3) # Shapiro-Wilk Test for normality
  ) %>%
  mutate(mean_sd = paste(mean, sd, sep = " \u00b1 "))  %>% # add a column containing mean and sd together, \u00b1 is the unicode for "+-"
  pivot_wider(names_from = Condition,
              # increase number of columns and decrease number of rows to make a nice summary table for presentation
              values_from = c(n, mean, sd, p_shapiro, mean_sd))
df_sum %>% 
  kable("html") %>%
  kable_styling() %>%
  scroll_box(height = "200px")
```

## Define the statistical tests that will be applied to the data  

```{r define stats functions}
## Define functions (levene/ t/ welch/ wilcox) that allow you to obtain the respective p-values
fun.levene.test <- function(x){
  a = leveneTest(Log2FoldChange ~ Condition, data = x)
  b = round(a[[3]][[1]], digits = 30)
}

fun.t.test <- function(x){
  a = t.test(Log2FoldChange ~ Condition, alternative = "two.sided", var.equal = TRUE, data = x)
  b = round(a[[3]], digits = 30)
}

fun.welch.test <- function(x){
  a = t.test(Log2FoldChange ~ Condition, alternative = "two.sided", var.equal = FALSE, data = x)
  b = round(a[[3]], digits = 30)
}

fun.wilcox.test <- function(x){
  a = wilcox.test(Log2FoldChange ~ Condition, exact=FALSE, data = x)
  b = round(a[[3]], digits = 30)
}
```


## Create a nested data frame and run statistical tests  

```{r nested df for stat analysis}
# Create a nested data frame that contains a list of lists for individual genes. Subsequently you will apply the test functions on each element (list) nested within the data frame.
df_nested <- df_stats%>% 
  group_by(Gene) %>% 
  nest()

# perform levene/ t/ welch/ wilcox testwith each element of the nested data frame
df_com <- df_nested %>% 
  mutate(
    levene = map_dbl(data, fun.levene.test), 
    t = map_dbl(data, fun.t.test),
    welch = map_dbl(data, fun.welch.test),
    wilcox = map_dbl(data, fun.wilcox.test)
  )

# combine with df_sum
df_sumcom <- df_sum %>% 
  left_join(df_com, by = "Gene")
```


## Define and run the decision tree  

All tests will be run on the data, however the comparison decision tree will select and highlight the appropriate test that meets the requirements (based onn the distribution of the data and variance).  

```{r decision tree}
# Create the comparison decision tree
pipe_com_choice <- . %>%
  mutate(shapiro = ifelse(p_shapiro_WT < 0.05 |
                            p_shapiro_KO < 0.05, "*", "n.s.")) %>%
  mutate(
    compare_which = case_when(
      shapiro != "n.s."                 ~ "wilcox",
      shapiro == "n.s." & levene > 0.05 ~ "t",
      shapiro == "n.s." & levene < 0.05 ~ "welch"
    )
  )  %>%
  mutate(
    test_p = ifelse(
      compare_which == "wilcox",
      wilcox,
      ifelse(compare_which == "t", t, welch)
    ),
    p_star = case_when(
      test_p < 0.0001 ~ "****",
      test_p < 0.001 ~ "***",
      test_p < 0.01 ~ "**",
      test_p < 0.05 ~ "*",
      test_p > 0.05 ~ "n.s."
    )
  )


# apply the decision tree
df_sumcom_choice <- df_sumcom %>% 
  pipe_com_choice()
```

## Multiple testing correction  
Perform multiple testing correction using the Benjamini & Hochberg (1995) method.  

```{r BH correction}
# calculate adjusted p values with the Benjamini & Hochberg (1995) method  ("BH" or its alias "fdr")
df_sumcom_choice <- df_sumcom_choice %>% 
  mutate(padj = p.adjust(test_p, method = "BH"), n = length(Gene)) %>% 
  mutate(padj_star = case_when(padj  < 0.0001 ~ "****",
                                padj  < 0.001 ~ "***",
                                padj < 0.01 ~ "**",
                                padj  < 0.05 ~ "*",
                                padj  > 0.05 ~ "n.s.")) 
```


## Create data frame with stats results  

The column "compare_which" indicates which test was selected with the decision tree, with "test_p" and "padj" indicating the associated test values.  

```{r stats final}
results_stats_final <- df_sumcom_choice %>% 
  select(!data) %>% 
  mutate(ID = ID, Details = Details) # add ID and Details info
# save to table
write.csv(results_stats_final, paste("../output/tables/results_stats_final_",ID,".csv", sep = ""))

results_stats_final %>% 
  kable("html") %>%
  kable_styling() %>%
  scroll_box(height = "200px")
```



### Create a dataframe for plotting the manually calculated p values  

This requires some cheating, since stat_pvalue_manual() (ggpubr) is usually used with predefined test functions. To use this function for adding the statistics to plots, we are creating a data frame that looks like the output of these functions but contains our p values and p star symbols.

```{r stat test for plotting}
# create a data frame that contains group1, group2. .y.and condition_+label
stat.test <- df_stats %>%
  group_by(Gene) %>%
  mutate(group1 = paste0("WT", "\n", "(N=", 6, ")")) %>% # create column containing WT x axis labels
  mutate(group2 = paste0("KO", "\n", "(N=", 6, ")")) %>% # create column containing KO x axis labels
  get_y_position(formula = Log2FoldChange ~ Condition, fun = "max") %>%  # calculate the y.position
  add_x_position() %>% # add x positions 1 and 2
  mutate(Condition = "KO") %>% # a Condition column is required for mapping of the statistics though it seems to not mater how it is filled.
  left_join(results_stats_final, stat.testx, by = "Gene") %>% # use left_join to add the results from the manually calculated statistics
  distinct(Gene, .keep_all = TRUE) # since the merging inflated the row number which now does not match the ggplot input, you need to select a subset.This subset contains all relevant details for each gene

stat.test %>% 
  kable("html") %>%
  kable_styling() %>%
  scroll_box(height = "200px")
```


# Plotting


## Set color and define a custom theme 

**Define colors used for plotting**  

Feel free to replace the Hex colors with color codes of your choosing.  

```{r set colors}
col <- c('#009ADC', '#FF1F5B')
```

**Define a custom theme/format to apply to the plots** 

Define a custom theme that can be applied to the plots. There are countless options that are not included and can be found online.  

```{r}

theme_custom <- theme(text = element_text(family = "", size = 14),
                      panel.background  = element_blank(),
                      title = element_text(color = "#000000"),
                      plot.title = element_text(hjust = 0.5),
                      legend.text = element_text(color = "#000000"),
                      legend.background = element_blank(),
                      legend.key = element_blank(),
                      axis.text = element_text(color = "#000000"),
                      axis.line = element_line(color = "#000000"),
                      strip.background = element_blank(),
                      panel.border = element_rect(colour = "#000000", fill=NA))
```


## Plot overview of all results

Plot of all results faceted by gene.  

```{r plot all genes,fig.height = 5, fig.width = 7, fig.align = "center"}
all_genes_plot <- df3 %>%
  ggplot(aes(x = Condition, y = Log2FoldChange, fill = Condition)) +
  geom_boxplot(alpha = 0.5, width = .5) + # add box plot
  scale_fill_manual(values = col) + # change colors
  geom_point( # add scatter plot
    shape = 21,
    size = 2,
    position = position_dodge(width = .75),
    color = "black",
    alpha = 0.5
  ) +
geom_hline(yintercept = 0, linetype = "dashed") + # add horizontal line indicating 0
  ggtitle(paste(ID)) + # add title based on ID
  facet_wrap( ~ Gene, nrow = 1) + # facet wrap by Gene
  stat_summary( # calculate and add mean expression value as black diamond
    fun = mean,
    colour = "black",
    geom = "point",
    shape = 18,
    size = 3,
    position = position_dodge(width = .75),
    show.legend = FALSE
  ) +
  stat_pvalue_manual(stat.test, label = 'padj_star', tip.length = 0.01, size = 7) + # add statistics
  theme_custom # apply custom theme formatting

# save plot
ggsave(
  "all_genes_plot.pdf",
  plot = all_genes_plot,
  device = "pdf",
  path = "../output/plots",
  height = 15,
  width = 20,
  scale = 1,
  units = "cm",
  dpi = 300,
  limitsize = TRUE,
)
# print to screen
print(all_genes_plot)
```



## Plot results for individual gene

The gene of interest is selected with the filter() function.  

**IMPORTANT**: Make sure to match your selection with the gene name in the stat_pvalue_manual() call that selects the fitting statistics results.  

"Condition_label" containing the N number is used instead of "Condition" for the X-axis labels here.

```{r plot selected gene, fig.height = 4, fig.width = 4, fig.align = "center"}
gene_1_plot <- df3 %>%
  filter(Gene == "gene_1") %>%
  ggplot(aes(x = Condition_label, y = Log2FoldChange, fill = Condition)) +
  geom_boxplot(alpha = 0.5, width = .5) +
  scale_fill_manual(values = col) + # change color to "col"
  geom_point(
    shape = 21,
    size = 2,
    position = position_dodge(width = .75),
    color = "black",
    alpha = 0.5
  ) +
  ggtitle("Gene_1") + # title
  xlab("Condition") +
  theme(
    plot.title = element_text(hjust = 0.5),
    # center title
    axis.text.x = element_text(
      size = 10,
      hjust = 0.5,
      vjust = 0.5
    ),
    axis.title = element_text(size = 15),
    legend.position = "none",
    strip.text = element_text(size = 10)
  ) + # set font size for facet label titles
  stat_summary(
    fun = mean,
    colour = "black",
    geom = "point",
    shape = 18,
    size = 3,
    position = position_dodge(width = .75),
    show.legend = FALSE
  ) +  # add the mean to the plots
  
  stat_pvalue_manual(filter(stat.test, Gene == "gene_1"), label = 'padj_star', size = 7) + # select appropriate 
    theme_custom # apply custom theme formatting


# save plot
ggsave(
  "gene_1_plot.pdf",
  plot = gene_1_plot,
  device = "pdf",
  path = "../output/plots",
  height = 10,
  width = 10,
  scale = 1,
  units = "cm",
  dpi = 300,
  limitsize = TRUE,
)
# print to screen
print(gene_1_plot)

```



```{r}
#sessionInfo()
```


